Scaling up the hyperparameter grid for Naomi

### 2023-03-22 src/dev_scale-grid/scale-grid.Rmd
PCA grid as mvQuad object, unit testing of PCA-AGHQ

```{r}
create_grid <- function(d, k) {
  gg <- mvQuad::createNIGrid(dim = d, type = "GHe", level = k)
  list(nodes = mvQuad::getNodes(gg), weights = as.numeric(mvQuad::getWeights(gg)))
}
```

@@

```{r}
s <- 5
d <- dim(E)[1]
gg_s <- create_grid(d = s, k = 2)
```

@@

```{r}
(n_nodes <- nrow(gg_s$nodes))
```

@@

```{r}
a_nodes <- gg_s$nodes
b_nodes <- t(E_s %*% diag(lambda[1:s]^{-0.5}) %*% t(a_nodes))

@@

```{r}
#' @param m Mean vector
#' @param C Covariance matrix
#' @param s Small grid dimension
#' @param k Number of points per small grid dimension
pca_rescale_grid <- function(m, C, s, k) {
  d <- nrow(C)
  stopifnot(d == length(m))
  eigenC <- eigen(C)
  lambda <- eigenC$values
  Lambda <- diag(lambda)
  E <- eigenC$vectors
  E_s <- E[, 1:s] 
  gg_s <- create_grid(d = s, k = k)
  a_nodes <- gg_s$nodes
  b_nodes <- t(E_s %*% diag(lambda[1:s]^{-0.5}, ncol = s) %*% t(a_nodes))
  for(j in 1:d) b_nodes[, j] <- b_nodes[, j] + m[j]
  b_weights <- gg_s$weights * sqrt(det(Lambda))
  return(list(nodes = b_nodes, weights = b_weights))
}
```

Check that this works as expected for a simple example:

@@

plot <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
  geom_contour(col = multi.utils::cbpalette()[1]) +
  coord_fixed(xlim = c(-2, 5), ylim = c(-2, 5), ratio = 1) +
  theme_minimal()

plot

test <- pca_rescale_grid(m = mu, C = cov, s = 1, k = 5)

plot +
  geom_point(
    data = data.frame(test$nodes),
    aes(x = X1, y = X2),
    col = multi.utils::cbpalette()[2],
    inherit.aes = FALSE
  )
```

Apply it to the Naomi grids:

```{r}
gg5 <- pca_rescale_grid(m = m, C = C, s = 5, k = 2)
gg7 <- pca_rescale_grid(m = m, C = C, s = 7, k = 2)
gg9 <- pca_rescale_grid(m = m, C = C, s = 9, k = 2)
```

>>>>

@@

```{r}
s <- 5
d <- dim(E)[1]
gg_s <- mvQuad::createNIGrid(dim = s, type = "GHe", level = 2) 
```

@@

```{r}
(n_nodes <- nrow(mvQuad::getNodes(gg_s)))
```

@@

```{r}
a_nodes <- mvQuad::getNodes(gg_s)
b_nodes <- t(E_s %*% diag(lambda[1:s]^{-0.5}) %*% t(a_nodes))

@@

```{r}
#' @param m Mean vector
#' @param C Covariance matrix
#' @param s Small grid dimension
#' @param k Number of points per small grid dimension
pca_rescale <- function(m, C, s, k) {
  d <- nrow(C)
  stopifnot(d == length(m))
  eigenC <- eigen(C)
  lambda <- eigenC$values
  Lambda <- diag(lambda)
  E <- eigenC$vectors
  E_s <- E[, 1:s] 
  gg_s <- mvQuad::createNIGrid(dim = s, type = "GHe", level = k) 
  nodes_out <- t(E_s %*% diag(lambda[1:s]^{-0.5}, ncol = s) %*% t(mvQuad::getNodes(gg_s)))
  for(j in 1:d) nodes_out[, j] <- nodes_out[, j] + m[j]
  weights_out <- mvQuad::getWeights(gg_s) * as.numeric(mvQuad::getWeights(mvQuad::createNIGrid(dim = d - s, type = "GHe", level = 1)))

  # Putting things into a mvQuad format manually
  gg <- mvQuad::createNIGrid(dim = d, type = "GHe", level = 1)
  gg$level <- rep(NA, times = d)
  gg$ndConstruction <- "PCA"
  gg$nodes <- nodes_out
  gg$weights <- weights_out
  return(gg)
}
```

### Testing with simple examples

Check that this works as visually expected for a simple example:

@@

plot <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
  geom_contour(col = multi.utils::cbpalette()[1]) +
  coord_fixed(xlim = c(-2, 4.5), ylim = c(-2, 4.5), ratio = 1) +
  labs(x = "", y = "") +
  theme_minimal()

gaussian_2d <- pca_rescale(m = mu, C = cov, s = 1, k = 5)

plot +
  geom_point(
    data = data.frame(mvQuad::getNodes(gaussian_2d)) %>%
      mutate(weight = mvQuad::getWeights(gaussian_2d)),
    aes(x = X1, y = X2, size = weight),
    col = multi.utils::cbpalette()[2],
    inherit.aes = FALSE
  ) +
  scale_size_continuous(limits = c(1, 2))
```

Test how this PCA grid works for some simple integrals.
We will use Alex's [unit tests](https://github.com/awstringer1/aghq/blob/master/tests/testthat/setup-01-optimization.R) from the `aghq` package.

```{r}
logfteta2d <- function(eta, y) {
  n <- length(y)
  n1 <- ceiling(n / 2)
  n2 <- floor(n / 2)
  y1 <- y[1:n1]
  y2 <- y[(n1 + 1):(n1 + n2)]
  eta1 <- eta[1]
  eta2 <- eta[2]
  sum(y1) * eta1 - (length(y1) + 1) * exp(eta1) - sum(lgamma(y1+1)) + eta1 +
    sum(y2) * eta2 - (length(y2) + 1) * exp(eta2) - sum(lgamma(y2+1)) + eta2
}

set.seed(84343124)
n1 <- 5
n2 <- 5
n <- n1 + n2
y1 <- rpois(n1, 5)
y2 <- rpois(n2, 5)

truemode2d <- c(log((sum(y1) + 1)/(length(y1) + 1)), log((sum(y2) + 1)/(length(y2) + 1)))

truelogint <- function(y) lgamma(1 + sum(y)) - (1 + sum(y)) * log(length(y) + 1) - sum(lgamma(y + 1))
truelognormconst2d <- truelogint(y1) + truelogint(y2)

objfunc2d <- function(x) logfteta2d(x, c(y1, y2))

funlist2d <- list(
  fn = objfunc2d,
  gr = function(x) numDeriv::grad(objfunc2d,x),
  he = function(x) numDeriv::hessian(objfunc2d,x)
)

opt_bfgs_2d <- aghq::optimize_theta(funlist2d, c(1.5, 1.5), control = default_control(method = "BFGS"))

norm_bfgs_2d_1 <- aghq::normalize_logpost(opt_bfgs_2d, 1, 1)
norm_bfgs_2d_3 <- aghq::normalize_logpost(opt_bfgs_2d, 3, 1)
norm_bfgs_2d_5 <- aghq::normalize_logpost(opt_bfgs_2d, 5, 1)
norm_bfgs_2d_7 <- aghq::normalize_logpost(opt_bfgs_2d, 7, 1)

pca_grid_bfgs_2d_1 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 1)
pca_grid_bfgs_2d_3 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 3)
pca_grid_bfgs_2d_5 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 5)
pca_grid_bfgs_2d_7 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 7)

plot_points <- function(gg) {
  gg$nodes %>%
    as.data.frame() %>%
    mutate(weights = gg$weights) %>%
    ggplot() +
    geom_point(
      aes(x = V1, y = V2, size = weights),
      alpha = 0.8,
      col = multi.utils::cbpalette()[2],
      inherit.aes = FALSE
    ) +
    scale_size_continuous(range = c(1, 2)) +
    labs(x = "", y = "", size = "Weight") +
    theme_minimal()
}

{plot_points(pca_grid_bfgs_2d_1) + plot_points(pca_grid_bfgs_2d_3)} / {plot_points(pca_grid_bfgs_2d_5) + plot_points(pca_grid_bfgs_2d_7)}

norm_pca_bfgs_2d_1 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_1, adapt = FALSE)
norm_pca_bfgs_2d_3 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_3, adapt = FALSE)
norm_pca_bfgs_2d_5 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_5, adapt = FALSE)
norm_pca_bfgs_2d_7 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_7, adapt = FALSE)

results_2d <- data.frame(
  "method" = c(
    "Truth",
    paste0("aghq, k = ", c(1, 3, 5, 7)),
    paste0("aghq-pca, k = ", c(1, 3, 5, 7))
  ),
  "lognormconst" = c(
    truelognormconst2d,
    norm_bfgs_2d_1$lognormconst, norm_bfgs_2d_3$lognormconst, norm_bfgs_2d_5$lognormconst, norm_bfgs_2d_7$lognormconst,
    norm_pca_bfgs_2d_1$lognormconst, norm_pca_bfgs_2d_3$lognormconst, norm_pca_bfgs_2d_5$lognormconst, norm_pca_bfgs_2d_7$lognormconst
  )
)

ggplot(results_2d, aes(x = method, y = lognormconst)) +
  geom_point(size = 3) +
  labs(x = "Method", y = "Log normalising constant") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 70, vjust = 0.5, hjust = 0.5))
```

### Application to Naomi

```{r}
gg3 <- pca_rescale(m = m, C = C, s = 3, k = 2)
nrow(mvQuad::getNodes(gg3))

gg5 <- pca_rescale(m = m, C = C, s = 5, k = 2)
nrow(mvQuad::getNodes(gg5))

gg7 <- pca_rescale(m = m, C = C, s = 7, k = 2)
nrow(mvQuad::getNodes(gg7))

gg9 <- pca_rescale(m = m, C = C, s = 9, k = 2)
nrow(mvQuad::getNodes(gg9))
```
